---
title: "MATH 3190 Lab 4"
author: "R Shiny (due 2/7/24)"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = FALSE, fig.width = 6, fig.height = 4)
options(width = 55)
library(tidyverse)
library(shiny)
```

Now its time to practice what we have learned in class and learn even more! Note that this lab should be completed in R Markdown or Quarto and Knitted to an html document. You will \`turn in' this homework by uploading to your GitHub Math_3190_Assignment repository in the "Labs/Lab_4" directory.

## Problem 1 - K Means Tutorial {.tabset}

This is a break down for the k-means example shiny app into easy to understand steps, adding a few shiny elements at a time. Is is adapted from the [Kmeans example code from Joe Cheng at Rstudio](https://shiny.rstudio.com/gallery/kmeans-example.html).

For each step, run the code and see how the app changes. Then below each code chunk, *describe exactly what is being changed/added at that step*. Make sure to install and load the `shiny` library before you get started.

### Step 00: Template

```{r, template, eval=F, echo=T}
ui <- fluidPage(
  
  # Put a titlePanel here
  
  
  sidebarLayout(
    # Sidebar. Put your inputs inside the sidebarPanel
    sidebarPanel(
      
    ),
    
    # Main panel. put your output plot here
    mainPanel(
      
    )
  )
)

server <- function(input, output, session) {

}

shinyApp(ui = ui, server = server)
```

**Response:** The `fluidpage()` function is stored as 'ui', this begins the app appearance and user-input (sort of initializing UX/UI of app). The shinyApp() function is the fundamental function to assemble UI and server-side logic into a single, launchable object. `Server logic` as a function takes`input`,`output`, and`session` as arguments to handle user input, generate the output, and manage the app's state. `sidebarLayout()` does what it sounds like: forming UX elements of the main panel, etc.

### Step 01: Title Panel

```{r, add title panel, eval=F, echo=T}
ui <- fluidPage(
  
  # Put a titlePanel here
  titlePanel("k-means clustering"),
  
  sidebarLayout(
    # Sidebar. Put your inputs inside the sidebarPanel
    sidebarPanel(
      
    ),
    
    # Main panel. put your output plot here
    mainPanel(
      
    )
  )
)

server <- function(input, output, session) {
  
}

shinyApp(ui = ui, server = server)
```

**Response:** The `titlePanel()` function takes a string argument, here "k-means clustering", and generates the necessary HTML code to render a title panel within the app's UI. Specifically, it renders the title within an `<h2>` heading tag by default, so that's really great. (It also would set the web browser's title bar to the specified `title` if `windowTitle` wasn't not provided).

### Step 02: UI inputs

```{r, add ui inputs, eval=F, echo=T}
ui <- fluidPage(
  
  # Put a titlePanel here
  titlePanel("k-means clustering"),
  
  sidebarLayout(
    # Sidebar. Put your inputs inside the sidebarPanel
    sidebarPanel(
      selectInput('xcol', 'X Variable', names(iris)),
      selectInput('ycol', 'Y Variable', names(iris),
                  selected=names(iris)[[2]]),
      numericInput('clusters', 'Cluster count', 3,
                   min = 1, max = 9)
    ),
    
    # Main panel. put your output plot here
    mainPanel(
      
    )
  )
)

server <- function(input, output, session) {
  
}

shinyApp(ui = ui, server = server)
```

**Response:** The `sidebarPanel()` object is provided arguments using the Fischer 'Iris' data set, in this case the `selectInput()` function creates drop-down 'selection' box in the UI for user choice of some defined options. For this app, the two menu objects created are for an $X$ variable and a $Y$ variable each from the the column names of the `iris` data set (the `choices` argument is *names(iris)* in this case). For the $Y$ variable, the initial selection is set to be the second column, so a default selection will provide a graph plotting two different variables. The `numericInput()` function creates a field for users to enter number values, and defines the interval, etc. for this input. In this case, in the context of K-means, the cluster sizes are defined for user-input parametrization.

### Step 03: UI outputs

```{r, add ui outputs, eval=F, echo=T}
ui <- fluidPage(
  
  # Put a titlePanel here
  titlePanel("k-means clustering"),
  
  sidebarLayout(
    # Sidebar. Put your inputs inside the sidebarPanel
    sidebarPanel(
      selectInput('xcol', 'X Variable', names(iris)),
      selectInput('ycol', 'Y Variable', names(iris),
                  selected=names(iris)[[2]]),
      numericInput('clusters', 'Cluster count', 3,
                   min = 1, max = 9)
    ),
    
    # Main panel. put your output plot here
    mainPanel(
      plotOutput('plot1')
    )
  )
)

server <- function(input, output, session) {
  
}

shinyApp(ui = ui, server = server)
```

**Response:** The \`plotOutput()\` function sets a 'placeholder' for the eventual plot to be rendered in the main panel, the content area, of the app's UI. It can be defined with width and height, say *width = "100%" , height = "500px",* etc. No UX effect takes place yet.

### Step 04: Output plot

```{r, add output plot, eval=F, echo=T}
ui <- fluidPage(
  
  # Put a titlePanel here
  titlePanel("k-means clustering"),
  
  sidebarLayout(
    # Sidebar. Put your inputs inside the sidebarPanel
    sidebarPanel(
      selectInput('xcol', 'X Variable', names(iris)),
      selectInput('ycol', 'Y Variable', names(iris),
                  selected=names(iris)[[2]]),
      numericInput('clusters', 'Cluster count', 3,
                   min = 1, max = 9)
    ),
    
    # Main panel. put your output plot here
    mainPanel(
      plotOutput('plot1')
    )
  )
)

server <- function(input, output, session) {
  output$plot1 <- renderPlot({
    plot(iris[,c("Sepal.Length","Sepal.Width")])
  })
}

shinyApp(ui = ui, server = server)
```

**Response:** The functional server logic is employed to render a plot. the output argument renders a first object by `renderPlot()` which creates a scatterplot of the specific sepal length and width arguments defined in a vector. This plot object is not dynamic/does not have a user interaction logic to it.

### Step 05: Reactive Axes

```{r, add reactive axes, eval=F, echo=T}
ui <- fluidPage(
  
  # Put a titlePanel here
  titlePanel("k-means clustering"),
  
  sidebarLayout(
    # Sidebar. Put your inputs inside the sidebarPanel
    sidebarPanel(
      selectInput('xcol', 'X Variable', names(iris)),
      selectInput('ycol', 'Y Variable', names(iris),
                  selected=names(iris)[[2]]),
      numericInput('clusters', 'Cluster count', 3,
                   min = 1, max = 9)
    ),
    
    # Main panel. put your output plot here
    mainPanel(
      plotOutput('plot1')
    )
  )
)

server <- function(input, output, session) {
  output$plot1 <- renderPlot({
    plot(iris[,c(input$xcol,input$ycol)])
  })
}

shinyApp(ui = ui, server = server)
```

**Response:** The `plot()` input of `renderPlot()` is updated so that the arguments pass the user input from the `selectInput` function. The plot becomes thus interactive. The `renderPlot()` function outputs scatterplots for pairs of numerical variables, and boxplots for the categorical variable of `species`.

### Step 06: K means

```{r, add K means clustering, eval=F, echo=T}
ui <- fluidPage(
  
  # Put a titlePanel here
  titlePanel("k-means clustering"),
  
  sidebarLayout(
    # Sidebar. Put your inputs inside the sidebarPanel
    sidebarPanel(
      selectInput('xcol', 'X Variable', names(iris)),
      selectInput('ycol', 'Y Variable', names(iris),
                  selected=names(iris)[[2]]),
      numericInput('clusters', 'Cluster count', 3,
                   min = 1, max = 9)
    ),
    
    # Main panel. put your output plot here
    mainPanel(
      plotOutput('plot1')
    )
  )
)

server <- function(input, output, session) {

  output$plot1 <- renderPlot({
    plot(iris[,c(input$xcol,input$ycol)],
         col = kmeans(iris[,c(input$xcol,input$ycol)], input$clusters)$cluster)
  })
}

shinyApp(ui = ui, server = server)
```

**Response:** The color argument of `plot()` is defined as the `$cluster` attribute obtained from the `kmeans()` algorithm executed from the user input for the iris data. The scatterplot will now be updated to represent points in colored clusters resulting from *k-means*. In this case, if the categorical `species` argument is selected an error message appears : '*NA/NaN/Inf in foreign function call (arg 1)*'.

### Step 07: Cluster centers

```{r, add cluster centers, eval=F, echo=T}
ui <- fluidPage(
  
  # Put a titlePanel here
  titlePanel("k-means clustering"),
  
  sidebarLayout(
    # Sidebar. Put your inputs inside the sidebarPanel
    sidebarPanel(
      selectInput('xcol', 'X Variable', names(iris)),
      selectInput('ycol', 'Y Variable', names(iris),
                  selected=names(iris)[[2]]),
      numericInput('clusters', 'Cluster count', 3,
                   min = 1, max = 9)
    ),
    
    # Main panel. put your output plot here
    mainPanel(
      plotOutput('plot1')
    )
  )
)

server <- function(input, output, session) {
  
  output$plot1 <- renderPlot({
    plot(iris[,c(input$xcol,input$ycol)],
         col = kmeans(iris[,c(input$xcol,input$ycol)], input$clusters)$cluster)
    points(kmeans(iris[,c(input$xcol,input$ycol)], input$clusters)$centers, pch = 4, cex = 4, lwd = 4)
  })
}

shinyApp(ui = ui, server = server)
```

**Response:** The `points()` function is added as an argument to `plot()` and takes the `$centers` attribute of the `kmeans()` function, again defined by user input for the iris data set. The formatting options of point size, etc. are also defined (here, large black X's). Categorical `species` selections continue to result in an error. It is also observed that `Sepal.Length` over `Petal.Length` seems to mis-apply the k-means cluster centers.

### Step 08: Save Kmeans

```{r, save k means result as reactive, eval=F, echo=T}
ui <- fluidPage(
  
  # Put a titlePanel here
  titlePanel("k-means clustering"),
  
  sidebarLayout(
    # Sidebar. Put your inputs inside the sidebarPanel
    sidebarPanel(
      selectInput('xcol', 'X Variable', names(iris)),
      selectInput('ycol', 'Y Variable', names(iris),
                  selected=names(iris)[[2]]),
      numericInput('clusters', 'Cluster count', 3,
                   min = 1, max = 9)
    ),
    
    # Main panel. put your output plot here
    mainPanel(
      plotOutput('plot1')
    )
  )
)

server <- function(input, output, session) {
  
  clusters <- reactive({
    kmeans(iris[,c(input$xcol,input$ycol)], input$clusters)
  })
  
  output$plot1 <- renderPlot({
    plot(iris[,c(input$xcol,input$ycol)],
         col = clusters()$cluster)
    points(clusters()$centers, pch = 4, cex = 4, lwd = 4)
  })
}

shinyApp(ui = ui, server = server)
```

**Response:** The shiny app object is made to be 'reactive'. That is, the `reactive()` function is used around the `kmeans()` function to define a variable called `clusters`. This is then used with parentheses like a function inside of the `renderPlot()` function. In using the app, a somewhat "off" appearing plot of cluster centers in plots of `Sepal.Length` over `Petal.Length` (and other similar combinations) from the previous iteration appear to be remedied.

### Step 09: Final App

```{r, final app, eval=F, echo=T}
ui <- fluidPage(
  
  # Put a titlePanel here
  titlePanel("k-means clustering"),
  
  sidebarLayout(
    # Sidebar. Put your inputs inside the sidebarPanel
    sidebarPanel(
      selectInput('xcol', 'X Variable', names(iris)),
      selectInput('ycol', 'Y Variable', names(iris),
                  selected=names(iris)[[2]]),
      numericInput('clusters', 'Cluster count', 3,
                   min = 1, max = 9)
    ),
    
    # Main panel. put your output plot here
    mainPanel(
      plotOutput('plot1')
    )
  )
)

server <- function(input, output, session) {
  
  selectedData <- reactive({
    iris[, c(input$xcol, input$ycol)]
  })
  
  clusters <- reactive({
    kmeans(selectedData(), input$clusters)
  })
  
  output$plot1 <- renderPlot({
    palette(c("#E41A1C", "#377EB8", "#4DAF4A", "#984EA3",
              "#FF7F00", "#FFFF33", "#A65628", "#F781BF", "#999999"))
    
    par(mar = c(5.1, 4.1, 0, 1))
    plot(selectedData(),
         col = clusters()$cluster,
         pch = 20, cex = 3)
    points(clusters()$centers, pch = 4, cex = 4, lwd = 4)
  })
}

shinyApp(ui = ui, server = server)
```

**Response:** In the final app version, the `reactive()` function is now used to define a variable called `selectedData` where user input is specifically passed for the iris data. `Clusters` reactive variable is then redefined by the `kmeans()` function using `selectedData` as its argument and the numerical `input$clusters` selection of the user. This is coded as such that these variables are than used in the `plot()` function inside of the `renderPlot()` function, creating a sort of 'piped' reactive object, were multiple arguments may be specified by the same variables and their attributes. A color palette is also defined with a vector argument and a set of margins. The error arising from the now defunct boxplot of categorical classes is not addressed, thus the app is specifically for the plotting of numerical iris data.

# Problem 2 - R Shiny for mtcars

The `mtcars` data set was extracted from the 1974 Motor Trend US magazine, and comprises fuel consumption and 10 aspects of automobile design and performance for 32 automobiles (1973--74 models). This data set is already built into **R**. You can load it by typing `data(mtcars)`. Once you do that, you should change the `cyl`, `vs`, `am`, `gear`, and `carb` variables to be factors rather than numeric.

Your job is to build an app that has the following:

-   A title for the app.
-   An input select box that lets the user specify which variable in the data set they are selecting.
-   An input select box that lets the user specify what type of plot to use, either a histogram, box plot, or bar plot.
-   The app should then output the plot of the variable that is selected. The plot should have a nice color, a title, and easy to read axes.
-   If the variable is quantitative and a bar plot is selected, then a message should appear to the user indicating the wrong type of plot is selected.
-   Likewise, if the variable is categorical and a histogram or box plot is selected, then a message should appear to the user indicating the wrong type of plot is selected.

Some hints:

-   The app will need a `ui` element, a `server` function, and the last line should be `shinyApp(server = server, ui = ui)`.
-   Add items to your app one at a time and run it often to make sure things are working as expected.
-   You'll need both a `renderText()` and a `renderPlot()` function in the server.
-   `if` statements are very useful in these shiny apps.
-   The variable selected by the user will be saved as an object in `input$`. You should use the `reactive()` functions to save it as an object in the `server` function and then access it in the `render...()` functions by typing its name with parentheses `()` at the end. For example, if you save it as `var_select` then you will access it as `var_select()` in the `reactive()` functions.
-   In `ggplot`, we are used to calling the name of the variable we are plotting in the `aes()` function. However, the variable we are plotting is being chosen by the user and is reactive. You can access the variable you want by putting `mtcars` as the first argument in `ggplot` and then by putting `aes(x = .data[[var_select()]])`, if you saved your variable as `var_select` in the server function. That `.data` will refer back to whatever data set you entered into the `ggplot` function. So, it will look like `ggplot(mtcars, aes(x = .data[[var_select()]]))`. Of course, you can change that `x` to an `y` where appropriate.

One more thing about R Markdown. If you change the **R** code chunk option from `eval = F` to `eval = T` after building your app, the code will run, but the app will not be interactive. It is just a static file. You can make it interaction by adding `runtime: shiny` (not indented) to the bottom of the YAML (under the `editor_options:` section). If you save the file, the Knit button at the top should change to a play button that says "Run Document". If you click that, it will make the document interactive. You cannot save and share the output of document in this state, though.

```{r mtcars_app, eval = T, echo = T}

data(mtcars)
mtcars <- mtcars |> 
  mutate(cyl = as.factor(cyl), 
         vs = as.factor(vs), 
         am = as.factor(am), 
         gear = as.factor(gear), 
         carb = as.factor(carb)
         ) |>
  tibble()

ui <- fluidPage(

  titlePanel("Motor Trend US - Car Data Viewer"),

  sidebarLayout(

    sidebarPanel(
      selectInput('x', 'Variable', names( mtcars )),
      
      selectInput('plot_Type', 'Plot Type: ', 
                  list(barplot = 'Barplot',
                       histogram = 'Histogram',
                       boxplot = 'Boxplot'))
    ),
    
    mainPanel(
    # output: user selected plot type or message representing mismatch of plot and data type
      
        plotOutput('thePlot'),
        
      tags$head(
        tags$style(
          '.top-text {
            position: absolute;
            top: 10%; 
            left: 50%;
            transform: translateX(-50%);
          }'
        )
      ),
        div(class = "top-text", h3(textOutput('plot_Mismatch'))
            )
    )
        
    
  )
    
)


server <- function(input, output, session) {
      
      selectedVar <- reactive({ mtcars[[ input$x ]] })

      is_Mismatch <- reactive({
        !is.numeric(selectedVar()) && 
          input$plot_Type %in% c('Histogram', 'Boxplot') ||
          is.numeric(selectedVar()) && input$plot_Type == 'Barplot'})

  output$thePlot <- renderPlot({
    
    if ( !is_Mismatch() ) {
      selectedPlotType <- switch(input$plot_Type,
                                 'Barplot' = geom_bar(color = 'sienna4', 
                                                      fill = "sienna3"),
                                 'Histogram' = geom_histogram(color = 'sienna4', 
                                                              fill = "sienna3"),
                                 'Boxplot' = geom_boxplot(color = 'sienna4', 
                                                          fill = "sienna3"))
      
      ggplot(data = mtcars, aes(x = .data[[input$x]], fill = selectedVar() ))+
        selectedPlotType+
        ggtitle(paste( input$plot_Type,'of', input$x ) )+
        theme(axis.title = element_text(size = 14),
          legend.title = element_text(size = 14),
          plot.title = element_text(size = 18, face = "bold"),
          panel.background = element_rect(fill = 'ivory', 
                                    color = 'navajowhite3', size=2,
                                    linetype = "solid")) 
    }
  })

  output$plot_Mismatch <- renderText({
    if (is_Mismatch()) {
      paste("You have selected ", input$plot_Type, 
            " which does not concur with variable data type. 
            Please make another selection.")
    }
  })

}

shinyApp(server = server, ui = ui)
      
```
